---
    title: Phoenix City Heat Island Analysis
    subtitle: Data Accessing
    author: Ousama Bin Zamir
    data: 2026-02-27
---

# Introduction

This document shows how the data was acquired. Python is being used for acquiring the data via GEE (Google Earth Engine) and R for census data via US Census Bureau API

# Initial 
```{python setup}
import sys
import ee
import geemap
import geopandas as gpd
import os
```

For authentication, the ee will allow you to config your Google Cloud project in browser and you will get a token. The token needs to be entered in the terminal manually.

```{python auth_init}
ee.Authenticate()
print("GEE Authentication Complete")
ee.Initialize()
```

# AOI and Dates Defining 

```{python AOI_dates}
# City of Phoenix coordinates
phoenix_coords = [
    [-112.469788, 33.192731],  
    [-112.469788, 33.986641],  
    [-111.816101, 33.986641],  
    [-111.816101, 33.192731], 
    [-112.469788, 33.192731]  
]
phx_bbox = ee.Geometry.Polygon(phoenix_coords)

# dates have been taken from first of May till the end of July
START_DATE = '2020-05-01'
END_DATE = '2020-07-31'
```

# Raster Data
## LST Data

For LST data the landsat data collection is being used.

```{python lst_data}
# so here the landsat 8 collection is being selected from Tier 1 Level 2.
landsat = ee.ImageCollection('LANDSAT/LC08/C02/T1_L2') \
    .filterBounds(phx_bbox) \
    .filterDate(START_DATE, END_DATE) \
    .filter(ee.Filter.lt('CLOUD_COVER', 20))

# So the data is ST_B10 (ST- Surface Temperatuer & B10 - band 10)
def calculate_lst(image):
    # converting to celsius
    thermal = image.select('ST_B10').multiply(0.00341802).add(149.0).subtract(273.15)
    return thermal

lst_collection = landsat.map(calculate_lst)
# taking the median values with the phoenix city bbox
lst_median = lst_collection.median().clip(phx_bbox)

# saving lst_median as geotiff with central Arizona CRS "Epsg:2223"
# the geotiff file is being automatically saved in your Google drive
export_lst = ee.batch.Export.image.toDrive(
    image=lst_median,
    description='PHOX_LST_Landsat8_Summer2020',
    folder='PHOX_Heat_Analysis',
    fileFormat='GeoTIFF',
    region=phx_bbox,
    scale=30,
    maxPixels=1e13,
    crs='EPSG:2223'
)
export_lst.start()
```

## NDVI Data

For vegetation index (NDVI) data the Copernicus, sentinel 2 data collection is being used with less than 20% cloud cover threshold.

```{python ndvi}
sentinel = ee.ImageCollection('COPERNICUS/S2_SR_HARMONIZED') \
    .filterBounds(phx_bbox) \
    .filterDate(START_DATE, END_DATE) \
    .filter(ee.Filter.lt('CLOUDY_PIXEL_PERCENTAGE', 20)) 

def calculate_ndvi(image):
    ndvi = image.normalizedDifference(['B8', 'B4']).rename('NDVI')
    return ndvi

ndvi_collection = sentinel.map(calculate_ndvi)
ndvi_median = ndvi_collection.median().clip(phx_bbox)

# saving vegetation index as geotiff with central Arizona CRS "Epsg:2223"
# the geotiff file is being automatically saved in your Google drive

export_ndvi = ee.batch.Export.image.toDrive(
    image=ndvi_median,
    description='PHOX_NDVI_sentinel_Summer2020',
    folder='PHOX_Heat_Analysis',
    fileFormat='GeoTIFF',
    region=phx_bbox,
    scale=10,
    maxPixels=1e13,
    crs='EPSG:2223'
)
export_ndvi.start()
```

# Census Data

## libraries

```{r setup}
library(tidycensus)
library(sf)
library(dplyr)
```

For Census Data first you need to register on US Census Bureau API site and after verification from their side you will recieve a API key.
```{r api_key}
census_api_key("api_key_here", install = TRUE, overwrite = TRUE)
# now this key will be automatically saved in your .Renviron file
```

We define all the variable we need and fetch the data
```{r census_data}
# necessary variable list 
variables_list = c(
    total_pop = "B01003_001",       # Total population
    median_income = "B19013_001",     # Median household income
    poverty_pop = "B17001_002",     # Population below poverty
    # lets add a fertility related variable as well
    births_past_year = "B13016_002" # Women who had a birth in past 12 months
    # lets add different age group
    paste0("B01001_", sprintf("%03d", c(
        1,    # this total population
        3:6,  # males under 5 till 15-17
        7:19,  # males from 18-19 age till 60-61
        20:25,  # males from 62-64 years old till 85+ age
        27:30, # females under 5 till 15-17
        31:43, # females from 18-19 age till 60-61
        44:49   # females from 62-64 years old till 85+ age 
    )))
  )

phoenix_census <- get_acs(
  geography = "tract",
  state = "AZ",
  county = "Maricopa",
  variables = variables_list,
  year = 2020,                          
  survey = "acs5",      # this is a 5-year estimates
  geometry = TRUE,                     
  output = "wide"                      
)

# lets mutate the data a according to our analysis need
phoenix_census <- phoenix_census %>%
    mutate(
        # both male and female under 18
        age_under_18 = rowSums(select(., ends_with(c("003E", "004E", "005E", "006E", 
        "027E", "028E", "029E", "030E")))),
        # population between 18 till 64
        age_18_64 = rowSums(select(., ends_with(paste0(sprintf("%03d", 7:19), "E")))) +
                    rowSums(select(., ends_with(paste0(sprintf("%03d", 31:43), "E")))),
        # population above 64 (elderly population)
        age_65_plus = rowSums(select(., ends_with(paste0(sprintf("%03d", 20:25), "E")))) +
                    rowSums(select(., ends_with(paste0(sprintf("%03d", 44:49), "E")))),
        # calculating percentages
        pct_poverty = (poverty_popE / total_popE) * 100,
        pct_under_18 = (age_under_18 / total_popE) * 100,
        pct_elderly = (age_65_plus / total_popE) * 100,
        # Crude birth rate per 1,000 population
        crude_birth_rate = (births_past_yearE / total_popE) * 1000,

        # verify it 
        total_check = age_under_18 + age_18_64 + age_65_plus
    )%>%
    select(GEOID, NAME, geometry, total_popE, median_incomeE,
         age_under_18, age_18_64, age_65_plus, pct_poverty,
         pct_under_18, pct_elderly, crude_birth_rate)

phoenix_census <- st_transform(phoenix_census, "EPSG:2223")

# Now we have the necessary data for the county Maricopa in which phoenix city is located 
```

